\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{ifthen,enumerate}
\usepackage{listings}
\usepackage{courier}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{pdfpages}
\usepackage{../thm}
\usepackage[left=3.8cm,top=3.8cm,right=3.8cm]{geometry}
\title{Responses for CS240 Assignment 3 Question 3}
\author{Alfred Chan -- 20392255}

\begin{document}
\lstset{
	language=C++,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{OliveGreen}\bfseries,
  stringstyle=\ttfamily\color{red!50!brown},
	numbers=left,
	showspaces=false,
	showstringspaces=false,
	stepnumber=1,
	basicstyle=\ttfamily,
	morekeywords={swap, and, then, initialize, push},
	frame=single,
	breaklines=true,
	tabsize=2
}

\maketitle
\begin{enumerate}[(a)]
\item Consider a modified version of $QuickSelect$ such that each result of the partition function is stored in a global variable linked list $L$. 
Another global variable, $num\_pop$, indicates the number of elements being popped.

Consider a partition function that partition an array into two parts:
\begin{itemize}
\item[left:] items that are greater or equal to the pivot element
\item[right:] items that are less than the pivot element
\end{itemize}
\begin{lstlisting}
partition(A[1..n], pivot) // start with element that is not popped
	i <- 1
	j <- n
	swap A[i] with A[pivot]
	while true
		while i <= n and A[i] >= A[1]
				i++
		while j >= 1 and A[j] < A[1]
				j--
		if i < j then break
		swap A[i] with A[j]
	swap A[i] with A[1]
	return i
\end{lstlisting}
The returned index is the i-th largest element in the array, and it is in its sorted position.

Now consider $QPQ$ as a modified version of $QuickSelect$.
Notice that the ``absolute'' position, resulting from partition, is being pushed into the linked list $L$:

\begin{lstlisting}
QPQ(1..n,k)
	p <- random number from 1 to n
	p <- partition(A[1..n], p)
	push (p + num_pop) into L // aka the ``absolute" position
	
	if k = p
		ret <- A
		A <- A + k // increment the address of A by k (k * size of data). This effectively "pops" k elements out of the array in constant time.
		num_pop += k
		return ret
	if k < p
		return QPQ(1..p-1, k)
	if k > p
		return QPQ(p+1..n, k)
\end{lstlisting}
\hfill $\blacksquare$

\item
The key idea: look for the upper and lower bound of indice in $L$ that are closest to 20. Then use them to bound $A$.

Let's call the function $A3Q3b$.
\begin{lstlisting}
A3Q3b(k <- 10) // k, by default 10
	big <- n - num_pop // size changed due to popping in QPQ
	small <- 1
	node_pt <- pointer to the head of L

	for each index in L
		index = index - num_pop // pointer arithmetics for retriving the relative position
		if index = k // the first k-th elements have being found (very lucky)
			ret <- A
			A <- A + k // pop k elements
			return ret
		else if index < k
			// small gets closer to k
			if small < index
				small <- index
		else // index > k
			// big gets closer to k
			if big < index
				big <- index

	// if nothing happened, the worst case is still QPQ(11..n,10);
	return QPQ(small..big, k);
\end{lstlisting}
The main reason that the above algorithm works is because whenever a partition is finish, the value of the returned index, $i$, is in the proper sorted place.
Hence, if the partition function is being called for a second time, $A[i]$ will not be affected because it is already in its sorted location. 
Also, the algorithm keeps track of the number of pops, which can be use to derieve the ``relative'' position.
Thus, even if $QPQ$ pops some elements, the indice in $L$ remain valid.
Lastly, $small$ and $big$ try to minimize the range for the second $QPQ$ call.
Hence, the algorithm can either be faster than or as fast as calling $QPQ(11..n,10)$.
\hfill $\blacksquare$

\item
Let's consider the expected running time for each part:
\begin{itemize}
\item[partition]
Notice that the array indice $i$ and $j$ go towards each other, and then the infinite while loop breaks whenever they pass each other.
So partition will always check the entire array.
In addition, the swap function is a constant operation. Therefore, the expected case for partition is $\Theta(n)$.
\item[QPQ]
The only difference between $QPQ$ and $QuickSelect$ are the lines where $A$ pops out elements, and where the result of partition is pushed into $L$.
Both operations are constant.
Now show that the expected case of $QPQ$ is $\Theta(n)$:

Consider the ``good'' and ``bad'' pivots from a sorted sequence of $A$ that is being split into 4 parts.
The values in between the second and the third parts are considered as ``good'' pivots, because these parts are close to the median, which allows $partition$ to eliminate about half of the elements.
Since the pivot is randomly chosen, the probability of selecting the second and the third is $\frac{1}{2}$, and selecting the rest is also $\frac{1}{2}$.

The worst case of a ``good'' pivot happens if the pivot is at the first or fourth quartile (25\% and 75\% of the data), so the worst case for ``good'' pivots is based on many $\frac{3}{4}$-split splits.
Also, the worst case of a ``bad'' pivot happens if the pivot belongs to the first or fourth parts of the sorted sequence, which $QPQ$ has to look for (almost) the entire list, hence $O(n)$.

Therefore, the worst case for $QPQ$ is

\begin{align*}
T(n) &= T(partition(n)) + \frac{1}{2}T(\text{bad pivots}) + \frac{1}{2}T(\text{good pivots})\\
	&= \Theta(n) + \frac{1}{2}n + \frac{1}{2}T(\frac{3}{4}n)\\
	&= \Theta(n) + \frac{1}{2}T(\frac{3}{4}n)\\
	&= \Theta(n) + \sum_{i = 0}^{\text{number of times n can be split by $\frac{3}{4}$}} (\frac{3}{4})^i\\
	\intertext{The following is true because $\sum_{i = 0}^{\infty} (\frac{3}{4})^i = 4$ (by geometric series)}\\
	&= \Theta(n) + O(1)\\
	&= \Theta(n)
\end{align*}

The expected best case is also $\Theta(n)$ because each element has to be compared at least once (otherwise there may be items that are greater/less/equal to the target element).
Hence, the expected time of $QPQ$ is bounded.
\end{itemize}
Thus, the expected time of $QPQ \in \Theta(n)$
\hfill $\blacksquare$

\item
Consider that each element in $A$ has a ``marker'' that distinguishes duplicated elements.
Then the probability of selecting a random index out of $n$ elements is $\frac{1}{n}$.
Consequently, the probability of selecting the 20-th element is $\frac{1}{n}$.
Also, the probabily of selecting pivots other than the 20-th element is $1-\frac{1}{n} = \frac{n-1}{n}$.

Consider $QPQ(1..n,10)$ is being called already.

Let $s \le n$ be the size of $L$. For the time that is used to look for small and big: 
it is either $\Theta(s)$ if 20 is not in $L$, since each element has to be compared; otherwise it is $O(s)$ (searching up to the size of $L$ until 20 has been found).
But the probability that an index is in $L$ is still based on $n$.

For executing $QPQ$, if the index 20 is found in $L$, then $A3Q3b$ returns the array address immediately.
Hence the time of execution of $QPQ$ is $0$.

Otherwise, $A3Q3b$ will call the second $QPQ$ with a possibly smaller bound, as explain near the pseudocode.
Though, the expected time of the second $QPQ$ call is still $\Theta(n)$ because $QPQ$'s expected time is $\Theta(n)$.

Then the expected time of $A3Q3b$ is (ignore the $\Theta$'s first)
\begin{align*}
T(A3Q3b) &= T(\text{searching small and big in L}) + T(\text{calling QPQ})\\
	&= \bigg( \frac{1}{n}T(\text{k in L}) + \frac{n-1}{n}T(\text{k not in L}) \bigg) + \bigg( \frac{1}{n}T(\text{k is in L}) + \frac{n-1}{n} T(\text{k not in L}) \bigg)\\
	&= \bigg( \frac{1}{n} O(s) + \frac{n-1}{n} s \bigg) + \frac{1}{n} \times 0 +  \frac{n-1}{n} n\\
	&= \bigg( O(1) + s-\frac{s}{n} \bigg) + \frac{n^2 -n}{n}\\
	&= \bigg( O(1) + s-\frac{s}{n} \bigg) + n - 1\\
	&= \bigg( n + s \bigg) + \bigg( O(1) + \frac{s}{n} - 1\bigg)\\
\intertext{Notice that $s$ is linear and $\frac{c}{n} \in o(1)$ for $c \le n$. Hence,}
	&= \Theta(n) + O(1)\\
	&= f(n) + O(1)
\end{align*}
As required.\\
Note: $T(\text{k in L})$ and $T(\text{k in L})$ are relative to either searching boundary or $QPQ$.
\hfill $\blacksquare$

\end{enumerate}
\end{document}

